Safe Transferable Regions

### 2015-11-20 paper/overview2.tex
Edits.
More small comments.

\paragraph{Transferable Regions} \name's \emph{transferable regions}
are an encapsulation of a data-structure that can be transferred
between autonomous entities (\eg, between two concurrently executing
threads or actors).  Hence, unlike stack regions, transferable regions
are not constrained to have a lexically scoped lifetime.  (Hence, we
also refer to them as \emph{dynamic} regions.)

Furthermore, transferable regions, unlike stack regions, are first
class values of \name: they are objects of the class \C{Region}, they
are created using the \C{new} keyword, and can be passed as arguments,
stored in data structures, and returned from methods.  A transferable
region is intended to encapsulate a single data-structure, consisting
of a collection of objects with a distinguished root object of some
type \C{T}, which we refer to as the region's \emph{root} object.  The
class \C{Region} is parametric over the type \C{T} of this root
object.

The \C{Region} constructor takes as a parameter a function that
constructs the root object: it creates a new region and invokes this
function, with the new region as the allocation context, to create the
root object of the region. The following code illustrates the
creation of a transferable region, whose root is an object of type
\C{A}.
% (\C{r}) of type \C{T}:
\begin{codejava} 
  Region<A> rgn = new Region<T>(() => new A())
\end{codejava} 
In the above code, \C{rgn} is called the \emph{handler} to the newly
created region, and is required to read the contents of the region, or
change its state. The class \C{Region} offers two methods: a \C{free}
method that deallocates the region (and all the objects allocated
within it), and \C{transfer} method that transfers the region to a
consumer process. It is an abstraction of two possible forms of
transfer: a transfer between two processes in a shared memory setting
or a transfer between two processes in a distributed, message-passing,
setting. The precise semantics of \C{transfer} are unimportant in the
context of the region type system and we will not discuss them
further.

@@

\paragraph{Open and Closed Regions} A transferable region must be
explicitly \emph{opened} using \name's \C{open} construct in order to
either read or update or allocate objects in the region.
Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
following: (a). It opens the transferable region handled by \C{rgn}
for allocation (i.e., makes it the current allocation context), (b).
binds the identifier \C{R} to this open region, and (c). initializes
the newly introduced local variable \C{v} to refer to the root object
of the region.  The \C{@R} part of the statement is optional and may
be omitted.  The \C{open} construct is intended to simplify the
problem of ensuring memory safety, as will be explained soon.  We
refer to a transferable region that has not been opened as a
\emph{closed} region.

\paragraph{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient).  Line 7 creates a new region to
store the output for time \C{t}, initializing it to contain an empty
list.  Line 9 opens the input region to process it.  Line 10 creates a
stack region \C{R0}.  Thus, the temporary objects created by the
iteration in line 11, for example, will be allocated in this stack
region that lives just long enough.  We open the desired output region
in line 12, so that the new output objects created by the invocation
of \C{selector} in line 13 are allocated in the output region.
Finally, the input region is freed in line 15. The output region at
\C{map[t]} stays as along as input messages with timestamp \C{t} keep
arriving. When the timing message for \C{t} arrives, the \C{onNotify}
method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
After the transfer, \C{SelectVertex} is no longer allowed to access
the \C{outRgn}, facilitating its deallocation.

@@

\subsection{Memory Safety} \label{sec:memory-safety}

Our goal is a type system that can ensure the memory safety of
programs that use the region constructs described above.  The key to
memory safety in \name is the following restriction: an object $o_1$
in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
similar restriction applies in the case where $o_1$ is a
stack-allocated variable.)

Enforcing this restriction is simple in the case of stack regions
since the outlives relation between stack regions can be inferred from
their lexical nesting. Unfortunately, inferring outlives relations
in presence of transferable regions is not easy.  \name imposes the following
protocol on the use of transferable regions to help simplify this
check.

A transferable region (that has not been freed or transferred) can be
in one of two possible states, \emph{open} or \emph{closed}. A newly
created region is in the closed state.  A region must be opened, using
the open construct (as explained previously), in order to read or
update or allocate an object within that region.  An open region
cannot be freed or transferred.  In particular, an open region is
guaranteed to be live for the entire duration of the open construct.
This allows the type system to infer a valid outlives relation between
the opened region and any stack region that is nested within the open
construct.

\begin{figure} 
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm} 
\vspace*{-0.25in} 
\end{figure}

The protocol for transferable regions is presented as a finite state
machine in Fig.~\ref{fig:region-fsm}.  
% A transferable region starts
% its lifetime in a \emph{closed} state, when it is created.
%
The safety of memory accesses in \name is now subject to the condition
that every transferable region correctly follows the state transition
discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
region type system statically guarantees the safety of all memory
accesses.
% In other words, the type system reduces the problem of ensuring
% memory safety in \name programs to the problem of enforcing the
% state transition discipline for transferable regions.

In \name, this enforcement is done at runtime by explicitly keeping
track of the \emph{current state} for $\RgnZ$ objects, and checking
the validity of every open, transfer, or free operation and throwing
an exception if it is invalid.  
% The challenge in enforcing this
% discipline statically is that transferable regions are first-class
% objects. Hence, the program can create multiple aliases for the same
% region, \eg, open it via one alias and free it via another.  Typestate
% verification in the presence of aliases is hard.  The checking can be
% done statically by preventing the creation of aliases using, \eg,
% linear types or unique types. However, this would be quite
% restrictive, in terms of expressiveness.
As explained previously, this is a reasonable trade-off in the context
of \name, as regions are coarse-grained objects, which are manipulated
infrequently, when compared to fine-grained objects that reside inside
these regions. Therefore, runtime overhead of checking the region's
state transition discipline is acceptable.

>>>>

\paragraph{Transferable Regions} 
% \name's \emph{transferable regions} are an encapsulation of a
% data-structure that can be transferred between autonomous entities
% (\eg, between two concurrently executing threads or actors).  Hence,
% unlike stack regions, transferable regions are not constrained to have
% a lexically scoped lifetime.  (Hence, we also refer to them as
% \emph{dynamic} regions.)

Transferable regions are first class values of \name: they are objects
of the class \C{Region}, they are created using the \C{new} keyword,
and can be passed as arguments, stored in data structures, and
returned from methods.  A transferable region is intended to
encapsulate a single data-structure, consisting of a collection of
objects with a distinguished root object of some type \C{T}, which we
refer to as the region's \emph{root} object.  The class \C{Region} is
parametric over the type \C{T} of this root object.

The \C{Region} constructor takes as a parameter a function that
constructs the root object: it creates a new region and invokes this
function, with the new region as the allocation context, to create the
root object of the region. The following code illustrates the
creation of a transferable region, whose root is an object of type
\C{A}.
% (\C{r}) of type \C{T}:
\begin{codejava} 
  Region<A> rgn = new Region<T>(() => new A())
\end{codejava} 
In the above code, \C{rgn} is called the \emph{handler} to the newly
created region, and is required to read the contents of the region, or
change its state. The class \C{Region} offers two methods: a \C{free}
method that deallocates the region (and all the objects allocated
within it), and \C{transfer} method that transfers the region to a
(possibly remote) consumer process. 
% It is an abstraction of two possible forms of
% transfer: a transfer between two processes in a shared memory setting
% or a transfer between two processes in a distributed, message-passing,
% setting. 
% The precise semantics of \C{transfer} are unimportant in the
% context of the region type system and we will not discuss them
% further. \dv{this is a bit counter-intuitive.}

@@

\paragraph{Open and Closed Regions} A transferable region must be
explicitly \emph{opened} using \name's \C{open} construct in order to
either read or update or allocate objects in the region.
Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
following: (a). It opens the transferable region handled by \C{rgn}
for allocation (i.e., makes it the current allocation context), (b).
binds the identifier \C{R} to this open region, and (c). initializes
the newly introduced local variable \C{v} to refer to the root object
of the region.  The \C{@R} part of the statement is optional and may
be omitted.  The \C{open} construct is intended to simplify the
problem of ensuring memory safety, as will be explained soon.  We
refer to a transferable region that has not been opened as a
\emph{closed} region. A transferable region can only be transferred or
freed when it is in closed state. The acceptable state transition
discipline over the lifetime of a transferable region is described
in Fig.~\ref{fig:region-fsm}. Enforcement of this discipline is done
at runtime.

\paragraph{Motivating Example}
% We now illustrate how the features of \name can be used to 
Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
\C{onReceive} method receives its input message in a
\emph{transferred} region (\ie, a \emph{closed} region whose ownership
is transferred to the recipient).  Line 7 creates a new region to
store the output for time \C{t}, initializing it to contain an empty
list.  Line 9 opens the input region to process it\footnote{We omit
\C{@R} annotation in \C{open} when we don't need \C{R}.}.  Line 10
creates a stack region \C{R0}.  Thus, the temporary objects created by
the iteration in line 11, for example, will be allocated in this stack
region that lives just long enough.  We open the desired output region
in line 12, so that the new output objects created by the invocation
of \C{selector} in line 13 are allocated in the output region.
Finally, the input region is freed in line 15. The output region at
\C{map[t]} stays as along as input messages with timestamp \C{t} keep
arriving. When the timing message for \C{t} arrives, the \C{onNotify}
method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
% After the transfer, \C{SelectVertex} is no longer allowed to access
% the \C{outRgn}, facilitating its deallocation.

@@

% \subsection{Memory Safety} \label{sec:memory-safety}

% Our goal is a type system that can ensure the memory safety of
% programs that use the region constructs described above.  The key to
% memory safety in \name is the following restriction: an object $o_1$
% in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
% a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
% similar restriction applies in the case where $o_1$ is a
% stack-allocated variable.)

% Enforcing this restriction is simple in the case of stack regions
% since the outlives relation between stack regions can be inferred from
% their lexical nesting. Unfortunately, inferring outlives relations
% in presence of transferable regions is not easy.  \name imposes the following
% protocol on the use of transferable regions to help simplify this
% check.

% A transferable region (that has not been freed or transferred) can be
% in one of two possible states, \emph{open} or \emph{closed}. A newly
% created region is in the closed state.  A region must be opened, using
% the open construct (as explained previously), in order to read or
% update or allocate an object within that region.  An open region
% cannot be freed or transferred.  In particular, an open region is
% guaranteed to be live for the entire duration of the open construct.
% This allows the type system to infer a valid outlives relation between
% the opened region and any stack region that is nested within the open
% construct.

\begin{figure} 
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm} 
\vspace*{-0.1in} 
\end{figure}

% The protocol for transferable regions is presented as a finite state
% machine in Fig.~\ref{fig:region-fsm}.  
% % A transferable region starts
% % its lifetime in a \emph{closed} state, when it is created.
% %
% The safety of memory accesses in \name is now subject to the condition
% that every transferable region correctly follows the state transition
% discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
% region type system statically guarantees the safety of all memory
% accesses.
% % In other words, the type system reduces the problem of ensuring
% % memory safety in \name programs to the problem of enforcing the
% % state transition discipline for transferable regions.

% In \name, this enforcement is done at runtime by explicitly keeping
% track of the \emph{current state} for $\RgnZ$ objects, and checking
% the validity of every open, transfer, or free operation and throwing
% an exception if it is invalid.  
% % The challenge in enforcing this
% % discipline statically is that transferable regions are first-class
% % objects. Hence, the program can create multiple aliases for the same
% % region, \eg, open it via one alias and free it via another.  Typestate
% % verification in the presence of aliases is hard.  The checking can be
% % done statically by preventing the creation of aliases using, \eg,
% % linear types or unique types. However, this would be quite
% % restrictive, in terms of expressiveness.
% As explained previously, this is a reasonable trade-off in the context
% of \name, as regions are coarse-grained objects, which are manipulated
% infrequently, when compared to fine-grained objects that reside inside
% these regions. Therefore, runtime overhead of checking the region's
% state transition discipline is acceptable.

### 2015-11-20 paper/type-inference.tex
Edits.

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an \C{A} object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the constructor and method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as $\rgn_0$ withroot p in
    letregion $\rgn_1$ in
      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}
% The first constraint is generated by $\typeOk$ on the type of
% $\C{p}$.  Second and third are generated by the the assignment
% expressions, and the last constraint is generated from the recursive
% call.

>>>>

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an \C{A} object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the constructor and method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as $p@\rgn_0$ in
    letregion $\rgn_1$ in
      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}
% The first constraint is generated by $\typeOk$ on the type of
% $\C{p}$.  Second and third are generated by the the assignment
% expressions, and the last constraint is generated from the recursive
% call.
